Precise spiking motifs in neurobiological and neuromorphic data

### 2022-11-11 manuscript.tex
synfire chain

\includegraphics[width=.7\textwidth]{figures/MainenSejnowski1995.png} % TODO replace by the simulation
\caption{The timing of the spikes produced following the repetition of a step stimuli is less reproducible than that to a noisy stimulus. While this seems paradoxical at first sight, this is due to the use of the same \emph{frozen} noise at each repetition and highlight the highly reproducible pattern of spikes when it is driven by a highly dynamic input. See this \href{https://github.com/laurentperrinet/2022_UE-neurosciences-computationnelles/blob/master/C_MainenSejnowski1995.ipynb}{notebook} for a replication of this result using a simple LIF model.}\label{fig:mainen}

@@

Theoretically, it was shown that a simple model may allow the propagation of such synfire chains~\citep{gewaltig_propagation_2001} by considering the dynamics of leaky integrate-and-fire (LIF) neurons in different groups of similar size. Each neuron of one group is connected by an excitatory synapse to the next. When a pulse is elicited in the first group, this may generate a spike in the next group. Depending on the concentration of synaptic weight values, this new activity may become more or less synchronized with respect to that of the previous pulse (as measured by the standard deviation of spike times within the pulse). Recursively applying this to a sequence of groups within a chain generates either a synfire propagation or not. A simple simulation of synfire propagation is shown in Figure~\ref{fig:diesman}. A crucial aspect of this emergence is in particular the dynamics of the spiking neuron model~\citep{gerstner_time_1995} and in particular the balance between excitation and inhibition~\citep{azouz_stimulus-selective_2008}. This balance was for instance modeled by feed-forward inhibition, a fine-scaled latency mechanism that is an essential ingredient in modelling so-called push-pull effects in the primary visual cortex~\citep{kremkow_push-pull_2016}. Further models have shown that such synfire chains could be embedded in topographies~\citep{aviel_embedding_2003}, while others used conductance-based neurons with feed-forward inhibition to improve the robustness of the propagation~\citep{kremkow_functional_2010}. In particular, this was implemented as a computational neuroscience benchmark model using the pyNN language~\citep{davison_pynn_2008} both in CPU-based and neuromorphic hardware~\citep{pfeil_six_2013}. % (see Figure~\ref{fig:pynn}).
%
%----------------------------%
%\begin{figure}
%\centering
%\includegraphics[width=.7\textwidth]{figures/fnins-07-00011-g004.jpg}
%\caption{A pyNN implementation of a Synfire chain with feedforward inhibition. The background noise  is implemented as random Gaussian current. The state space gives %the duration of synfire chains as a function of the protocol's parameters.}\label{fig:pynn}
%\end{figure}
%----------------------------%
%----------------------------%
\begin{figure}
\centering
\includegraphics[width=.4\textwidth]{figures/Diesmann_et_al_1999.png}
\caption{Simulation of a synfire propagation \href{https://brian2.readthedocs.io/en/stable/examples/frompapers.Diesmann_et_al_1999.html}{using Brian}. The model consists of 10 groups of 100 neurons each. A pulse with a given jitter is generated in the first group, which here generates a pulse after a certain processing delay in the second group with a lower jitter. This allows the propagation of the synchronous activity along the chain of the neural groups.}\label{fig:diesman}
\end{figure}
%----------------------------%

>>>>

\includegraphics[width=\textwidth]{figures/replicating_MainenSejnowski1995.pdf} 
\caption{The timing of the spikes produced following the repetition of a step stimuli is less reproducible than that to a noisy stimulus. The stimulus is shown in the top row, while trial repetitions of a simulated Leaky Integrate-and-Fire neuron are shown in the second row (with membrane potential represented by  dark blue color when light yellow colors when depolarized). While this seems paradoxical at first sight, this is due to the use of the same \emph{frozen} noise at each repetition and highlight the highly reproducible pattern of spikes when it is driven by a highly dynamic input. See this \href{https://github.com/laurentperrinet/2022_UE-neurosciences-computationnelles/blob/master/C_MainenSejnowski1995.ipynb}{notebook} for a replication of this result using a simple LIF model.}\label{fig:mainen}

@@

Theoretically, it was shown that a simple model may allow the propagation of such synfire chains~\citep{gewaltig_propagation_2001} by considering the dynamics of leaky integrate-and-fire (LIF) neurons in different groups of similar size. Each neuron of one group is connected by an excitatory synapse to the next. When a pulse is elicited in the first group, this may generate a spike in the next group. Depending on the concentration of synaptic weight values, this new activity may become more or less synchronized with respect to that of the previous pulse (as measured by the standard deviation of spike times within the pulse). Recursively applying this to a sequence of groups within a chain generates either a synfire propagation or not. A simple simulation of synfire propagation is shown in Figure~\ref{fig:diesman}. A crucial aspect of this emergence is in particular the dynamics of the spiking neuron model~\citep{gerstner_time_1995} and in particular the balance between excitation and inhibition~\citep{azouz_stimulus-selective_2008}. This balance was for instance modeled by feed-forward inhibition, a fine-scaled latency mechanism that is an essential ingredient in modelling so-called push-pull effects in the primary visual cortex~\citep{kremkow_push-pull_2016}. Further models have shown that such synfire chains could be embedded in topographies~\citep{aviel_embedding_2003}, while others used conductance-based neurons with feed-forward inhibition to improve the robustness of the propagation~\citep{kremkow_functional_2010}. In particular, this was implemented as a computational neuroscience benchmark model using the pyNN language~\citep{davison_pynn_2008} both in CPU-based and neuromorphic hardware~\citep{pfeil_six_2013}. % (see Figure~\ref{fig:pynn}).
%
%----------------------------%
%\begin{figure}
%\centering
%\includegraphics[width=.7\textwidth]{figures/fnins-07-00011-g004.jpg}
%\caption{A pyNN implementation of a Synfire chain with feedforward inhibition. The background noise  is implemented as random Gaussian current. The state space gives %the duration of synfire chains as a function of the protocol's parameters.}\label{fig:pynn}
%\end{figure}
%----------------------------%
%----------------------------%
\begin{figure}
\centering
\includegraphics[width=\textwidth]{figures/Diesmann_et_al_1999.pdf}
\caption{Simulation of a synfire propagation \href{https://brian2.readthedocs.io/en/stable/examples/frompapers.Diesmann_et_al_1999.html}{using Brian}. The model consists of 10 groups of 100 neurons each. A pulse with a given jitter is generated in the first group, which here generates a pulse after a certain processing delay in the second group with a lower jitter. This allows the propagation of the synchronous activity along the chain of the neural groups.}\label{fig:diesman}
\end{figure}
%----------------------------%
